---
title: "BioMarin Meetup: reshaping, transforming, and joining data in R (exercises)"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: tango
    theme: lumen
    df_print: paged
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
# create data folder
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE)
```

# Objectives 

This document outlines some common data reshaping tasks with `tidyr`. 

The slides for this presentation are [here](https://mjfrigaard.github.io/data-transformations-joins/Index.html#1)

There is also an accompanying RStudio.Cloud project 



Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) and [`tidyr`]() on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

## Load packages 

`tidyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```


## Manging data {.tabset}

We're going to use some of the `fs` package functions for managing our files and data. 

### Create folders  

```{r fs-fir-info}
# fs::dir_create("")
```

### Folder trees 


### File paths

## 